---
title: "digits_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{digits_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SparseBayesICA)
```

The first thing we need to do is create some example data. The following function
will generate some "brain maps" that are a single slice. The activation regions for
each component will correspond to a single digit from 0--9. Any covariate effects 
will be generated as sub-parts of the activation regions. This will be made more
clear in plots shortly.

```{r}
simulation_truth <- generate_digits_example_data(N = 15, Q = 5,
                                       n_continuous_cov = 1, n_discrete_cov = 0,
                                       slice_width = 75)
```

This function has written output nifti files to your local package-specific 
data directory. To make this easy to find, it was been returned as a part of the
`simulation_truth` object above. We can take a look at the folder contents:

```{r}
data_directory <- simulation_truth$data_directory
list.files(data_directory)
```

Before moving on, let's take a look at the data. The true values for the 
population level maps and the covariate effect maps are stored in the `simulation_truth` 
object. We can use ggplot to plot them:

```{r}

```
